This notebook runs through the code required to produce Figure 4. 

First, we generate simulated data. 

```{r, warning=FALSE, message=FALSE}
library(MASS)
library(Matrix)
library(reshape2)
library(ggplot2)
library(ggthemes)
library(cowplot)
library(glmmTMB)
library(scales)
library(viridis)
library(tidyr)
library(pbkrtest)
library(lmerTest)
library(edgeR)
library(miloR)
```

```{r}
initializeFullZsim <- function(Z, cluster_levels, stand.cols=FALSE){
    # construct the full Z with all random effect levels
    n.cols <- ncol(Z)
    col.classes <- apply(Z, 2, class)
    i.z.list <- list()
    for(i in seq_len(n.cols)){
        i.class <- col.classes[i]
        if(i.class %in% c("factor")){ # treat as factors
            i.levels <- levels(Z[, i, drop=FALSE])
            i.levels <- as.factor(paste(sort(as.integer(i.levels))))
            i.z <- sapply(i.levels, FUN=function(X) (Z[, i] == X) + 0, simplify=TRUE)
        } else if(i.class %in% c("character")){
            i.levels <- unique(Z[, i, drop=FALSE])
            i.levels <- as.factor(paste(sort(as.integer(i.levels))))
            i.z <- sapply(i.levels, FUN=function(X) (Z[, i] == X) + 0, simplify=TRUE)
        } else if(i.class %in% c("numeric")){ # split into unique levels if integer levels
            i.mod <- all(Z[, i, drop=FALSE] %% 1 == 0)
            if(isTRUE(i.mod)){
                i.levels <- unique(Z[, i])
                i.levels <- as.factor(paste(sort(as.integer(i.levels))))
                i.z <- sapply(i.levels, FUN=function(X) (Z[, i] == X) + 0, simplify=TRUE)
            } else{
                i.z <- Z[, i, drop=FALSE] # if float then treat as continuous
            }
        } else if(i.class %in% c("integer")){
            i.levels <- (unique(Z[, i]))
            i.levels <- as.factor(paste(sort(as.integer(i.levels))))
            i.z <- sapply(i.levels, FUN=function(X) (Z[, i] == X) + 0, simplify=TRUE)
        }
        colnames(i.z) <- cluster_levels[[colnames(Z)[i]]]
        
        # to standardise or not?
        if(isTRUE(stand.cols)){
            q <- ncol(i.z)
            i.ident <- diag(1L, nrow=nrow(i.z), ncol=nrow(i.z))
            i.star <- i.z - ((i.ident %*% i.z)/q)
            i.z <- i.star
        }
        
        i.z.list[[colnames(Z)[i]]] <- i.z
    }
    full.Z <- do.call(cbind, i.z.list)
    return(full.Z)
}
```


```{r}
SimulateXZ <- function(N, n.fe, n.re, re.levels, fe.levels){
    
    # create a per-level mean effect for each FE
    if(length(fe.levels) != n.fe){
        stop("List entries need to match number of input fixed effects")
    }
    
    if(length(re.levels) != n.re){
        stop("List entries need to match number of input random effects")
    }
    
    # create the design matrices
    X <- matrix(0L, ncol=n.fe+1, nrow=N)
    X[, 1] <- 1
    colnames(X) <- c("Intercept", names(fe.levels))
    
    Z <- matrix(0L, ncol=n.re, nrow=N)
    
    for(i in seq_len(n.fe)){
        if(fe.levels[[i]] == 1){
            X[, i+1] <- sapply(seq_len(N), FUN=function(B){
                rnorm(1, mean=0, sd=1)
                })
        } else if(fe.levels[[i]] == 2){
            X[, i+1] <- sapply(seq_len(N), FUN=function(B){
                sample(c(0, 1), 1)
                })
            X[, i+1] <- as.factor(X[, i+1])
        }else{
            X[, i+1] <- sapply(seq_len(N), FUN=function(B){
                sample(seq_len(fe.levels[[i]]), 1)
                })
            X[, i+1] <- as.factor(X[, i+1])
        }
    }
    
    # Make categorical effects 0 or 1 (not 1 or 2)
    X[,2] <- X[,2] - 1
    
    for(j in seq_len(n.re)){
        if(re.levels[[j]] == 1){
            Z[, j] <- sapply(seq_len, FUN=function(R){
                rnorm(1, mean=1, sd=1)
            })
        } else{
            Z[, j] <- sapply(seq_len(N), FUN=function(R){
                sample(seq_len(re.levels[[j]]), 1)
            })
            Z[, j] <- factor(Z[, j], levels=c(1:re.levels[[j]]))
        }
    }
    colnames(Z) <- names(re.levels)

    sim.data <- do.call(cbind.data.frame, list(X, Z))
    return(sim.data)
}


SimulateY <- function(N, X, Z, fe.betas, re.sigmas,
                           dispersion, grand.mean, n.fe, n.re,
                           re.levels,
                           fe.levels){
    
    # create a per-level mean effect for each FE
    if(length(fe.levels) != n.fe){
        stop("List entries need to match number of input fixed effects")
    }
    
    if(length(re.levels) != n.re){
        stop("List entries need to match number of input random effects")
    }
    
    # construct the full Z
    random.levels <- sapply(seq_len(length(re.levels)), FUN=function(RX) {
        rx.name <- names(re.levels)[RX]
        paste(rx.name, seq_len(re.levels[[rx.name]]), sep="_")
        }, simplify=FALSE)
    names(random.levels) <- names(re.levels)

    full.Z <- initializeFullZsim(Z, random.levels)
    
    # get a combination over random effects 
    # and sample each level from the same ~Normal(0, sigma)
    # note that the variance would be G if we also had random slopes
    re.thetas <- list()
    for(i in seq_len(length(re.levels))){
        i.re <- names(random.levels[i])
        i.levels <- length(random.levels[[i.re]])
        i.re.means <- rnorm(n=i.levels, 0, sd=sqrt(re.sigmas[[i.re]])) # sample a random effect value
        i.re.list <- sapply(seq_len(i.levels), FUN=function(X) i.re.means[X])
        names(i.re.list) <- random.levels[[i.re]]
        re.thetas[[i.re]] <- i.re.list
    }
    
    B <- full.Z %*% unlist(re.thetas)
    # map the fixed effects to mean values
    betas <- c(grand.mean, unlist(fe.betas))
    Beta <- X %*% betas

    i.error <- matrix(data = rnorm(N, mean=0, sd=0.001), ncol = 1)
    
    # construct the y.means equation, depending on desired distribution and FE/RE
    y.means <- exp(Beta + B) 
    y.means <- y.means + i.error
    
    y.counts <- rnbinom(N, mu = y.means, size = dispersion)

    sim.data <- data.frame("Mean.Count"=y.counts)
    sim.data <- do.call(cbind.data.frame, list(sim.data, X, Z))

    return(sim.data)
}
```

Create a simulation with known parameter values to benchmark glmm vs TMB.

```{r, warning=FALSE, message=FALSE}
set.seed(5)
N=500
fe.levels <- list("FE1"=2)
re.levels <- list("RE1"=10)
design.sim <- SimulateXZ(N=N, n.fe=length(fe.levels), n.re=length(re.levels), re.levels=re.levels, fe.levels=fe.levels)
j <- 1
sim.list <- c()

for (i in seq(0.1, 1, 0.2)){
    r.dispersion <- 2
    fe.betas=list("FE1"=i)
    re.sigmas=list("RE1"=0.05)
    grand.mean=2
    sim.list[[j]] <- SimulateY(N=N, X=sapply(design.sim[,1:2], as.numeric), Z=design.sim[,3,drop=FALSE],
                        fe.betas=fe.betas, re.sigmas=re.sigmas, dispersion=r.dispersion,
                        grand.mean=grand.mean, n.fe=length(fe.betas), n.re=length(re.sigmas),
                        re.levels=re.levels, fe.levels=fe.levels)
    j <- j + 1

}
names(sim.list) <- seq(0.1, 1, 0.2)

glmmWrapperTMB <- function(sim.df){
    glmm.tmb <- glmmTMB(Mean.Count ~ FE1 + (1|RE1), sim.df, family=nbinom2(link="log"), REML = TRUE)
    var <- VarCorr(glmm.tmb)
    se <- sqrt(diag(vcov(glmm.tmb,full=TRUE)))
    out.list <- list("coefficients"=c(glmm.tmb[["fit"]]$parfull[1:2]),
                     "sigma"=unlist(var$cond),
                     "se"=unname(se[1:2]))
}
a <- system.time(glmm.sim.listTMB <- lapply(sim.list, glmmWrapperTMB))

glmmWrapperREML <- function(sim.df){
    z.model <- sapply(sim.df[,4, drop = FALSE], as.numeric)
    x.model <- sapply(sim.df[,2:3], as.numeric)
    y.model <- sapply(sim.df[,1, drop=FALSE], as.numeric)
    rand.levels <- lapply(seq_along(colnames(z.model)), FUN=function(X) unique(z.model[, X]))
    names(rand.levels) <- colnames(z.model)

    model.list <- fitGLMM(X=x.model, Z=z.model, y=y.model, random.levels=rand.levels, REML = TRUE, 
                    dispersion=r.dispersion, glmm.control=list(theta.tol=1e-6, max.iter=20))

    out.list <- list("coefficients"=c(model.list$FE, model.list$RE),
                         "sigma"=model.list$Sigma, 
                     "se"=model.list$SE)
}
b <- system.time(glmm.sim.listREML <- lapply(sim.list, glmmWrapperREML))

glmmWrapperPQL <- function(sim.df){
    glmm.pql <- glmmPQL(Mean.Count ~ FE1, random = ~ 1 | RE1, data = sim.df, family=nbinom2(link="log"))
    var <- VarCorr(glmm.pql)
    se <- sqrt(diag(vcov(glmm.pql,full=TRUE)))
    out.list <- list("coefficients"=glmm.pql$coefficients$fixed,
                     "sigma"=as.numeric(var[1,1]),
                     "se"=unname(se[1:2]))
}
c <- system.time(glmm.sim.listPQL <- lapply(sim.list, glmmWrapperPQL))

glmmWrapperGLMER <- function(sim.df){
    glmer1 <- glmer(Mean.Count ~ FE1 + (1|RE1), data = sim.df, family=nbinom2(link="log"))
    var <- VarCorr(glmer1)
    se <- sqrt(diag(vcov(glmer1,full=TRUE)))
    out.list <- list("coefficients"=glmer1@beta,
                     "sigma"=as.numeric(var$RE1),
                     "se"=unname(se[1:2]))
}
d <- system.time(glmm.sim.listGLMER <- lapply(sim.list, glmmWrapperGLMER))
```

Plot glmm vs TMB for B1 slope estimates and standard error estimates. 

```{r}
coeff <- lapply(glmm.sim.listTMB, `[[`, 1)
sigma <- lapply(glmm.sim.listTMB, `[[`, 2)
se <- lapply(glmm.sim.listTMB, `[[`, 3)
coeff_mat <- t(data.frame(matrix(unlist(coeff), nrow=2)))[,1:2]
se_mat <- t(data.frame(matrix(unlist(se), nrow=2)))
coeff_mat <- data.frame(cbind(unlist(sigma), se_mat, coeff_mat))
coeff_mat$ground_truth <- as.numeric(names(glmm.sim.listTMB))
colnames(coeff_mat) <- c("sigma", "se.int", "se.fe", "int", "fe", "ground_truth")
coeff_mat$analysis <- "nb_glmm_tmb"
coeff_pivoted <- pivot_longer(coeff_mat, cols = 1:5)

coeff <- lapply(glmm.sim.listREML, `[[`, 1)
sigma <- lapply(glmm.sim.listREML, `[[`, 2)
se <- lapply(glmm.sim.listREML, `[[`, 3)
coeff_mat <- t(data.frame(matrix(unlist(coeff), nrow=12)))[,1:2]
se_mat <- t(data.frame(matrix(unlist(se), nrow=2)))
coeff_mat <- data.frame(cbind(unlist(sigma), se_mat, coeff_mat))
coeff_mat$ground_truth <- as.numeric(names(glmm.sim.listTMB))
colnames(coeff_mat) <- c("sigma", "se.int", "se.fe", "int", "fe", "ground_truth")
coeff_mat$analysis <- "nb_glmm_REML"
coeff_pivoted2 <- pivot_longer(coeff_mat, cols = 1:5)

coeff <- lapply(glmm.sim.listPQL, `[[`, 1)
sigma <- lapply(glmm.sim.listPQL, `[[`, 2)
se <- lapply(glmm.sim.listPQL, `[[`, 3)
coeff_mat <- t(data.frame(matrix(unlist(coeff), nrow=2)))[,1:2]
se_mat <- t(data.frame(matrix(unlist(se), nrow=2)))
coeff_mat <- data.frame(cbind(unlist(sigma), se_mat, coeff_mat))
coeff_mat$ground_truth <- as.numeric(names(glmm.sim.listTMB))
colnames(coeff_mat) <- c("sigma", "se.int", "se.fe", "int", "fe", "ground_truth")
coeff_mat$analysis <- "nb_glmm_PQL"
coeff_pivoted3 <- pivot_longer(coeff_mat, cols = 1:5)

coeff <- lapply(glmm.sim.listGLMER, `[[`, 1)
sigma <- lapply(glmm.sim.listGLMER, `[[`, 2)
se <- lapply(glmm.sim.listGLMER, `[[`, 3)
coeff_mat <- t(data.frame(matrix(unlist(coeff), nrow=2)))[,1:2]
se_mat <- t(data.frame(matrix(unlist(se), nrow=2)))
coeff_mat <- data.frame(cbind(unlist(sigma), se_mat, coeff_mat))
coeff_mat$ground_truth <- as.numeric(names(glmm.sim.listTMB))
colnames(coeff_mat) <- c("sigma", "se.int", "se.fe", "int", "fe", "ground_truth")
coeff_mat$analysis <- "nb_glmm_GLMER"
coeff_pivoted4 <- pivot_longer(coeff_mat, cols = 1:5)

to_plot <- rbind.data.frame(coeff_pivoted, coeff_pivoted2, coeff_pivoted3, coeff_pivoted4)
to_plot_f <- dplyr::filter(to_plot, name =="fe")
to_plot_s <- dplyr::filter(to_plot, name =="se.fe")

ggplot(to_plot_f, aes(y = value, x = as.factor(ground_truth), group = analysis)) +
    geom_bar(width = 0.7, stat = "identity", color = "black", position = position_dodge(), aes(fill=analysis)) +
    theme_bw() +
    xlab("Ground truth \u03B21") +
    ylab(expression(hat("\u03B21"))) +
    theme(legend.position = "none", axis.text.x=element_text(colour="black"), axis.text.y=element_text(colour="black"), text = element_text(size = 18)) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0.1, 1, 0.2), expand = c(0,0)) +
    scale_fill_manual(values = c("#A78DB4", "#BEBEBE", "#8DA0CB", "#83A98A"), name="Model", labels = c("lme4:glmer", "MASS::glmmPQL", "miloR::NB-GLMM", "glmmTMB"))

ggsave("fig4a.tiff", path = "/Users/kluzer01/Desktop/figures/", height = 4, width = 4.5, device='tiff', dpi=700)

ggplot(to_plot_s, aes(y = value, x = as.factor(ground_truth), group = analysis)) +
    geom_bar(width = 0.7, stat = "identity", color = "black", position = position_dodge(), aes(fill=analysis)) +
    theme_bw() +
    xlab("Ground truth \u03B21") +
    ylab(expression(widehat("\u03B21 SE"))) +
    theme(legend.position = "top", axis.text.x=element_text(colour="black"), axis.text.y=element_text(colour="black")) +
    theme(text = element_text(size = 18)) +
    scale_y_continuous(limits = c(0, 0.08), breaks = seq(0, 0.07, 0.02), expand = c(0,0)) +
    scale_fill_manual(values = c("#A78DB4", "#BEBEBE", "#8DA0CB", "#83A98A"), name="Model", labels = c("lme4:glmer", "MASS::glmmPQL", "miloR::NB-GLMM", "glmmTMB"))

ggsave("fig4b.tiff", path = "/Users/kluzer01/Desktop/figures/", height = 4, width = 10, device='tiff', dpi=700)
```

Next, we want to benchmark the pvalues obtained with the Satterthwaite method. We need to run a version of the glmm algorithm that outputs the final y* values -- these are normally distributed y values.

Run the following code:

```{r}
runGLMM <- function(X, Z, y, random.levels=NULL, REML=TRUE,
                    glmm.control=list(theta.tol=1e-6, max.iter=100),
                    dispersion = NULL){

    # model components
    # X - fixed effects model matrix
    # Z - random effects model matrix
    # y - observed phenotype

    theta.conv <- glmm.control[["theta.tol"]] # convergence for the parameters
    max.hit <- glmm.control[["max.iter"]]

    # OLS for the betas is usually a good starting point for NR
    curr_beta <- solve((t(X) %*% X)) %*% t(X) %*% log(y + 1)

    # create full Z with expanded random effect levels
    full.Z <- initializeFullZ(Z, cluster_levels = random.levels)
    colnames(full.Z) <- unlist(random.levels)

    # sample random value for RE us
    curr_u <- Matrix(runif(ncol(full.Z), 0, 1), ncol=1, sparse = TRUE)

    # sample variances of the us
    curr_sigma <- Matrix(runif(ncol(Z), 0, 1), ncol=1, sparse = TRUE)
    rownames(curr_sigma) <- colnames(Z)

    u_indices <- sapply(seq_along(random.levels),
                        FUN=function(RX) which(colnames(full.Z) %in% random.levels[[RX]]),
                        simplify=FALSE)

    #create a single variable for the thetas
    curr_theta <- do.call(rbind, list(curr_beta, curr_u))

    #compute mu.vec using inverse link function
    mu.vec <- exp((X %*% curr_beta) + (full.Z %*% curr_u))

    theta_diff <- rep(Inf, nrow(curr_theta))
    sigma_diff <- Inf

    #compute variance-covariance matrix G
    curr_G <- initialiseG(cluster_levels=random.levels, sigmas=curr_sigma)
    G_inv <- computeInv(curr_G)

    conv.list <- list()
    iters <- 1
    meet.conditions <- !((all(theta_diff < theta.conv)) & (sigma_diff < theta.conv) | iters >= max.hit)

    while(meet.conditions){
        #compute all matrices - information about them found within their respective functions
        D <- computeD(mu=mu.vec)
        D_inv <- solve(D)
        y_star <- computey_star(X=X, curr_beta = curr_beta, full.Z = full.Z, D_inv = D_inv, curr_u = curr_u, y=y)
        V <- computeV(mu=mu.vec, r=dispersion)
        W <- computeW(D_inv=D_inv, V=V)
        W_inv <- solve(W)
        V_star <- computeV_star(full.Z=full.Z, curr_G=curr_G, W=W)
        V_star_inv <- solve(V_star)
        V_partial <- computeV_partial(full.Z=full.Z, random.levels=random.levels, u_indices=u_indices)

        matrix.list <- preComputeMatrices(V_star_inv, V_partial, X, curr_beta, full.Z, curr_u, y_star)
        #---- First estimate variance components with Newton Raphson procedure ---#
        if (isFALSE(REML)) {
            P <- computeP_REML(V_star_inv=V_star_inv, X=X) #needed for computeVarCovar
            PV <- computePV(V_partial=V_partial, P=P) #needed for computeVarCovar
            score_sigma <- sigmaScore(matrix_list=matrix.list, V_star_inv=V_star_inv, V_partial=V_partial, random.levels=random.levels)
            information_sigma <- sigmaInformation(V_star_inv=V_star_inv, V_partial=V_partial, random.levels=random.levels)
        } else if (isTRUE(REML)) {
            P <- computeP_REML(V_star_inv=V_star_inv, X=X)
            PV <- computePV(V_partial=V_partial, P=P)
            score_sigma <- sigmaScoreREML(PV=PV, P=P, y_star=y_star, random.levels=random.levels)
            information_sigma <- sigmaInformationREML(PV=PV, random.levels=random.levels)
        }
        sigma_update <- FisherScore(score_vec=score_sigma, hess_mat=information_sigma, theta_hat=curr_sigma)
        sigma_diff <- abs(sigma_update - curr_sigma)

        # update sigma, G, and G_inv
        curr_sigma <- sigma_update
        curr_G <- initialiseG(cluster_levels=random.levels, sigmas=curr_sigma)
        G_inv <- solve(curr_G)

        #---- Next, solve pseudo-likelihood GLMM equations to compute solutions for B and u---####
        theta_update <- solve_equations(X=X, W_inv=W_inv, full.Z=full.Z, G_inv=G_inv, curr_beta=curr_beta, curr_u=curr_u, y_star=y_star)
        if (isTRUE(theta_update)) {
          stop("Hessian is computationally singular - cannot solve GLMM")
        }

        theta_diff <- abs(theta_update - curr_theta)

        # update B, u and mu_vec to determine new values of score and hessian matrices
        curr_theta <- theta_update
        rownames(curr_theta) <- c(colnames(X), colnames(full.Z))
        curr_beta <- curr_theta[colnames(X), , drop=FALSE]
        curr_u <- curr_theta[colnames(full.Z), , drop=FALSE]
        mu.vec <- exp((X %*% curr_beta) + (full.Z %*% curr_u))

        if (any(is.infinite(mu.vec))) {
          stop("Estimates increasing to infinity - cannot solve GLMM.")
        }

        iters <- iters + 1
        meet.conditions <- !((all(theta_diff < theta.conv)) & (all((sigma_diff) < theta.conv))| iters >= max.hit)
    }

    SE <- calculateSE(X=X, full.Z=full.Z, W_inv=W_inv, G_inv=G_inv)
    Zscore <- calculateZscore(curr_beta=curr_beta, SE=SE)
    Va <- computeVarCovar(random.levels, PV)
    mint <- nrow(curr_beta)
    cint <- nrow(curr_u)
    coeff.matrix <- makeCoefMatrix(X=X, full.Z=full.Z, W_inv=W_inv, G_inv=G_inv)
    df <- Satterthwaite_df(coeff.mat=coeff.matrix, mint=mint, cint=cint, SE=SE, V_a=Va,
                           V_partial=V_partial, G_inv=G_inv, curr_sigma=curr_sigma, curr_beta=curr_beta, random.levels=random.levels)
    Pvalue <- computePvalue(Zscore=Zscore, df=df)

    converged <- ((all(theta_diff < theta.conv)) & (all(abs(sigma_diff) < theta.conv)))
    final.list <- list("FE"=as.vector(curr_beta),
                       "RE"=as.vector(curr_u),
                       "Sigma"=as.vector(curr_sigma),
                       "Theta.Converged"=theta_diff < theta.conv,
                       "Sigma.Converged"=sigma_diff < theta.conv,
                       "converged"=converged,
                       "Iters"=iters,
                       "Dispersion"=dispersion,
                       "SE"=SE,
                       "Zscore"=Zscore,
                       "df" = df,
                       "G"=curr_G,
                       "VSTAR"=V_star,
                       "Hessian"=information_sigma,
                       "pvalue"=Pvalue, 
                       "y_star"=y_star)

    return(final.list)
}

computeW <- function(D_inv, V){
    W = D_inv %*% V %*% D_inv
    return(W)
}

computeV <- function(mu, r){
    # compute diagonal matrix of variances
    v.vec <- ((mu**2/r)) + mu
    V <- Matrix(0L, ncol=length(mu), nrow=length(mu), sparse = TRUE)
    diag(V) <- v.vec
    return(V)
}

computeD <- function(mu){
    # D is diag(mu_i)
    D <- Matrix(0L, ncol=length(mu), nrow=length(mu), sparse = TRUE)
    diag(D) <- mu
    return(D)
}

computeV_star <- function(full.Z, curr_G, W){
    # V_star_R <- (full.Z %*% curr_G %*% t(full.Z)) + W
    V_star_C <- computeVStar(as.matrix(full.Z), as.matrix(curr_G), as.matrix(W))
    return(V_star_C)
}

computey_star <- function(X, curr_beta, full.Z, D_inv, curr_u, y){
    y_star <- ((X %*% curr_beta) + (full.Z %*% curr_u)) + D_inv %*% (y - exp((X %*% curr_beta) + (full.Z %*% curr_u)))
    return(y_star)
}

computeV_partial <- function(full.Z, random.levels, u_indices){
    # wrapper for c++ function
    # currently doesn't support a sparse matrix (why???)
    V_partial_vec_C <- pseudovarPartial(x=as.matrix(full.Z), rlevels=random.levels, cnames=colnames(full.Z))
    return(V_partial_vec_C)
}

computeVStar <- function(Z, G, W) {
    .Call('_miloR_computeVStar', PACKAGE = 'miloR', Z, G, W)
}

computeVstar_inverse <- function(full.Z, curr_G, W_inv){
    # compute the inverse of V_star using Henderson-adjusted Woodbury formula, equation (18)
    # (A + UBU^T)^-1 = A^-1 - A^-1UB[I + U^TA^-1UB]^-1U^TA^-1
    # Only requires A^-1, where B = ZGZ^T, A=W, U=Z
    vsinv_C <- invertPseudoVar(as.matrix(W_inv), as.matrix(curr_G), as.matrix(full.Z))
    return(vsinv_C)
}

preComputeMatrices <- function(V_star_inv, V_partial, X, curr_beta, full.Z, curr_u, y_star){
    # precompute certain matrices from matrix multiplications that are needed > once
    mat.list <- list()
    mat.list[["XBETA"]] <- X %*% curr_beta
    mat.list[["ZU"]] <- full.Z %*% curr_u

    mat.list[["YSTARMINXB"]] <- y_star - mat.list[["XBETA"]]
    mat.list[["XTVSTAR"]] <- t(X) %*% V_star_inv
    mat.list[["VSTARX"]] <- V_star_inv %*% X

    return(mat.list)
}

sigmaScore <- function(matrix_list, V_partial, V_star_inv, random.levels){
    score_vec <- NA
    for (i in seq_along(random.levels)) {
        LHS <- -0.5*matrix.trace(V_star_inv %*% V_partial[[i]])
        rhs.1 <- t(matrix_list[["YSTARMINXB"]]) %*% V_star_inv %*% V_partial[[i]]
        rhs.2 <- rhs.1 %*% V_star_inv
        RHS <- 0.5* rhs.2 %*% (matrix_list[["YSTARMINXB"]])
        score_vec[i] <- LHS + RHS
    }
    return(score_vec)
}

sigmaInformation <- function(V_star_inv, V_partial, random.levels) {

    sigma_info <- Matrix(0L, ncol=length(V_partial), nrow=length(V_partial))

    for(i in seq_along(V_partial)){
        for(j in seq_along(V_partial)){
            inner.1 <- V_star_inv %*% V_partial[[i]]
            inner.2 <- inner.1 %*% V_star_inv
            sigma_info[i, j] <- 0.5*matrix.trace(inner.2 %*% V_partial[[j]])
        }
    }
    return(sigma_info)
}

sigmaScoreREML <- function(y_star, PV, P, random.levels){
    score_vec <- Matrix(0L, ncol=1, nrow=length(random.levels), sparse=FALSE)
    for (i in seq_along(random.levels)) {
        LHS <- -0.5 * matrix.trace(PV[[i]])
        rhs.1 <- t(y_star) %*% PV[[i]]
        rhs.2 <- rhs.1 %*% P
        RHS <- 0.5 * (rhs.2 %*% y_star)
        score_vec[i, ] <- LHS + RHS
    }
    return(score_vec)
}

sigmaInformationREML <- function(PV, random.levels) {
    # this should be a matrix
    sigma_info <- Matrix(0L, ncol=length(random.levels), nrow=length(random.levels))

    for(i in seq_along(random.levels)){
        for(j in seq_along(random.levels)){
            sigma_info[i, j] <- 0.5*matrix.trace(crossprod(PV[[i]], PV[[j]]))
        }
    }

    return(sigma_info)
}

computeP_REML <- function(V_star_inv, X) {
    # breaking these down to individual steps speeds up the operations considerably
    tx.m <- t(X) %*% V_star_inv
    x.inv <- computeInv(tx.m %*% X)
    vx <- V_star_inv %*% X
    Pminus <- vx %*% x.inv
    tx.inv <- t(X) %*% V_star_inv
    P <- V_star_inv - Pminus %*% tx.inv
    return(P)
}

FisherScore <- function(score_vec, hess_mat, theta_hat, lambda=1e-5, det.tol=1e-10, cond.tol=1e-15){
    # sequentially update the parameter using the Newton-Raphson algorithm
    # theta ~= theta_hat + hess^-1 * score
    # this needs to be in a direction of descent towards a minimum

    theta_new <- tryCatch({
        theta_hat + solve(hess_mat) %*% score_vec
    }, error=function(cond){
        message("Hessian is singular. Original error message:")
        error(cond)
        return(NULL)
    }, finally={

    })
    rownames(theta_new) <- rownames(theta_hat) # not sure why these get stripped off during computation
    return(theta_new)
}

initialiseG <- function(cluster_levels, sigmas){
    # construct the correct size of G given the random effects and variance components
    # names of cluster_levels and columns of Z must match
    # the independent sigmas go on the diagonal and the off-diagonal are the crossed/interactions
    # sigmas must be named
    sum.levels <- sum(unlist(lapply(cluster_levels, length)))
    G <- sparseMatrix(i=sum.levels, j=sum.levels, repr="C", x=0L)
    dimnames(G) <- list(unlist(cluster_levels), unlist(cluster_levels))
    i <- j <- 1

    for(x in seq_len(nrow(sigmas))){
        x.q <- length(cluster_levels[[rownames(sigmas)[x]]])
        diag(G[c(i:(i+x.q-1)), c(i:(i+x.q-1)), drop=FALSE]) <- sigmas[x, ] # is this sufficient to transform the sigma to the model scale?
        i <- j <- i+x.q
    }
    return(as.matrix(G))
}

initializeFullZ <- function(Z, cluster_levels, stand.cols=FALSE){
    # construct the full Z with all random effect levels
    n.cols <- ncol(Z)
    col.classes <- apply(Z, 2, class)
    i.z.list <- list()
    for(i in seq_len(n.cols)){
        i.class <- col.classes[i]
        if(i.class %in% c("factor")){ # treat as factors
            i.levels <- levels(Z[, i, drop=FALSE])
            i.levels <- as.factor(paste(sort(as.integer(i.levels))))
            i.z <- sapply(i.levels, FUN=function(X) (Z[, i] == X) + 0, simplify=TRUE)
        } else if(i.class %in% c("character")){
            i.levels <- as.factor(unique(Z[, i, drop=FALSE])) # ordering is arbitrary
            i.z <- sapply(i.levels, FUN=function(X) (Z[, i] == X) + 0, simplify=TRUE)
        } else if(i.class %in% c("numeric")){ # split into unique levels if integer levels
            i.mod <- all(Z[, i, drop=FALSE] %% 1 == 0)
            if(isTRUE(i.mod)){
                i.levels <- unique(Z[, i])
                i.levels <- as.factor(paste(sort(as.integer(i.levels))))
                i.z <- sapply(i.levels, FUN=function(X) (Z[, i] == X) + 0, simplify=TRUE)
            } else{
                i.z <- Z[, i, drop=FALSE] # if float then treat as continuous
            }
        } else if(i.class %in% c("integer")){
            i.levels <- (unique(Z[, i]))
            i.levels <- as.factor(paste(sort(as.integer(i.levels))))
            i.z <- sapply(i.levels, FUN=function(X) (Z[, i] == X) + 0, simplify=TRUE)
        }

        colnames(i.z) <- cluster_levels[[colnames(Z)[i]]]

        # to standardise or not?
        if(isTRUE(stand.cols)){
            q <- ncol(i.z)
            i.ident <- diag(1L, nrow=nrow(i.z), ncol=nrow(i.z))
            i.star <- i.z - ((i.ident %*% i.z)/q)
            i.z <- i.star
        }

        i.z.list[[colnames(Z)[i]]] <- i.z
    }
    full.Z <- do.call(cbind, i.z.list)
    # full.Z <- Matrix(full.Z, sparse = FALSE)
    return(full.Z)
}

computePV <- function(V_partial, P){
  PV <- list()
  for (i in 1:length(V_partial)) {
    PV[[i]] <- P %*% V_partial[[i]]
  }
  return(PV)
}

solve_equations <- function(X, W_inv, full.Z, G_inv, curr_beta, curr_u, y_star){
  
    UpperLeft <- t(X) %*% W_inv %*% X
    UpperRight <- t(X) %*% W_inv %*% full.Z
    LowerLeft <- t(full.Z) %*% W_inv %*% X
    LowerRight <- t(full.Z) %*% W_inv %*% full.Z + G_inv

    LHS <- rbind(cbind(UpperLeft, UpperRight), cbind(LowerLeft, LowerRight))
    RHS <- rbind((t(X) %*% W_inv %*% y_star), (t(full.Z) %*% W_inv %*% y_star))

    theta_update <- tryCatch({solve(LHS) %*% RHS}
             , error = function(e){
               exit <- TRUE
               }
    )
    return(theta_update)
}

mapUtoIndiv <- function(full.Z, curr_u, random.levels){
    # map the vector of random effects to the full nx1 vector
    rand.levels <- names(random.levels)
    indiv.u.list <- list()

    for(j in seq_along(rand.levels)){
        j.G <- matrix(0L, ncol=nrow(full.Z), nrow=nrow(full.Z))
        j.re <- rand.levels[j]
        j.levels <- random.levels[[j.re]]
        j.b <- full.Z[, j.levels] %*% curr_u[j.levels, ]
        indiv.u.list[[j.re]] <- j.b
    }

    return(indiv.u.list)
}

computeInv <- function(x){
    # Compute x^-1 from x
    # need to check that x is not singular - use tryCatch - if matrix is singular then report error message
    x_inv <- tryCatch(expr={
        solve(x)
    },
    error=function(cond){
        message("Matrix cannot be inverted - most likely singular")
        message(cond)
        return(NULL)
    },
    finally={
    })
    return(x_inv)
}

pseudovarPartial <- function(x, rlevels, cnames) {
    .Call('_miloR_pseudovarPartial', PACKAGE = 'miloR', x, rlevels, cnames)
}
```


```{r}
# side note, checking satterthwaite
dispersion <- 2
sim.df <- sim.list[[1]]

# run glmm
X <- sapply(design.sim[,1:2], as.numeric)
Z <- design.sim[,3, drop = F]
random.levels <- list("RE1"=paste("RE1", levels(as.factor(design.sim$RE1)), sep="_"))
y <- sim.df$Mean.Count

nb.glm <- glmmTMB(Mean.Count ~ 1 + FE1 + (1|RE1), data=sim.df, family=nbinom2(link="log"), REML=TRUE, se=TRUE)
#summary(nb.glm)
model.list <- runGLMM(X=X, Z=Z, y=y, random.levels=random.levels, REML = TRUE, dispersion=dispersion, glmm.control=list(theta.tol=1e-6, max.iter=50))

sim.df$ystar <- as.vector(model.list$y_star)
test <- lmerTest::lmer(ystar ~ 1 + FE1 + (1|RE1), data = sim.df, REML = TRUE)
#summary(test)

coef(summary(test))
cbind(model.list$FE, model.list$SE, model.list$df, model.list$Zscore, model.list$pvalue)
coef(summary(nb.glm))

plot <- cbind.data.frame(matrix(coef(summary(test))[,5]), model.list$pvalue, matrix(coef(summary(nb.glm))$cond[,4]))
colnames(plot) <- c("lmer", "glmm", "TMB")
plot$variable <- rownames(plot)
plot_longer <- pivot_longer(plot, cols = c("lmer", "glmm", "TMB"))
plot_longer$method <- rep(c("Satterthwaite", "Satterthwaite", "ANOVA"), 2)
plot_longer1 <- dplyr::filter(plot_longer, variable == "Intercept")

laber <- c('FE1'=paste0("\U03B2","1"), 'Intercept'=paste0("\U03B2","0"))
ggplot(plot_longer, aes(y = -log10(value), x = name)) +
    facet_wrap(~variable, scales = "free", ncol = 1, labeller = as_labeller(laber)) +
    geom_bar(width = 0.8, stat = "identity", color = "black", position = position_dodge(), aes(fill=method)) +
    theme_bw() + 
    xlab("Model") +
    ylab("-log(*P*-value)") +
    theme(text = element_text(size = 18)) +
    scale_y_continuous(expand = c(0, 0, 0.2, 0)) +
    theme(legend.position = "none", axis.text.x=element_text(colour="black"), axis.text.y=element_text(colour="black"), axis.title.y = ggtext::element_markdown()) +
    scale_fill_manual(values = c("#FFDBA9", "#B0D8F2"), name="Degrees of Freedom")

ggsave("fig4d.tiff", path = "/Users/kluzer01/Desktop/figures/", height = 7, width = 4.5, device='tiff', dpi=700)

```
