---
title: "testing_speed"
output: html_document
---

```{r}
library(miloR)
library(SingleCellExperiment)
library(scater)
library(dplyr)
library(patchwork)
library(tidyr)
library(tibble)
```

Here we import and plot elapsed time, memory and k value analyses with 200,000 cell dataset.

To start with, we plot the elapsed time. To generate the data, run 'benchmark_speed.R' using 'benchmark_speed.sh'

```{r}
time_df <- read.csv("/Users/kluzer01/Documents/milo_analysis_2022/Figure_2/time_df.csv")
time.df2 <- pivot_longer(data.frame(time_df), cols = c(reduced_dim, graph), names_to = "method")
ggplot(time.df2, aes(x=(n_cells), y=value/60, group_by = method, fill = method)) +
    theme_bw(base_size=16) +
    geom_point(shape=21, size=3) +
    scale_fill_manual(values = c("#8da0cb", "#fc8d62"), name="Sampling scheme", labels = c("Graph", "Reduced dim")) +
    labs(x="Number of cells", y="Elapsed time (min)") +
    theme(legend.position = "top", axis.text.x=element_text(colour="black"), axis.text.y=element_text(colour="black")) +
    geom_text(aes(label=round(value/60, 1)),hjust=-0.3,vjust=0.5) +
    scale_x_continuous(limits = c(30000, 210000), labels = c("50,000", "100,000", "150,000", "200,000"), breaks = c(50000, 100000, 150000, 200000))

ggsave("fig2alegen.tiff", path = "/Users/kluzer01/Desktop/figures/", height = 4, width = 7, device='tiff', dpi=700)
```

Next, plot memory analysis with 200,000 cells. To generate the data, run 'memory_sim.R' using 'benchmark_speed.sh' (make sure to change the file name from benchmark_speed.R to memory_sim.R).

```{r}
mem_df <- read.csv("/Users/kluzer01/Documents/milo_analysis_2022/Figure_2/mem.df")
mem_df2 <- pivot_longer(data.frame(mem_df), cols = c(reduced_dim, graph), names_to = "method")

ggplot(mem_df2, aes(x=n_cells, y=value, group_by = method, fill = method)) +
    theme_bw(base_size=16) +
    geom_point(shape=21, size=3) +
    scale_fill_manual(values = c("#8da0cb", "#fc8d62"), name="Sampling scheme", labels = c("Graph", "Reduced dim")) +
    labs(x="Number of cells", y="Maximum memory (MB)") +
    theme(legend.position = "none", axis.text.x=element_text(colour="black"), axis.text.y=element_text(colour="black")) +
    geom_text(aes(label=round(value, 0)),hjust=1.3,vjust=0.5) +
    scale_x_continuous(limits = c(30000, 210000), labels = c("50,000", "100,000", "150,000", "200,000"), breaks = c(50000, 100000, 150000, 200000))

ggsave("fig2b.tiff", path = "/Users/kluzer01/Desktop/figures/", height = 4, width = 4.5, device='tiff', dpi=700)
```


Next, plot 200,000 cells with different values of k: 10, 30, 50, 70.
To generate the data, run 'benchmark_speed.R' (bottom section that is currently commented out) using 'benchmark_speed.sh' (again, remember to change file name that you are running).

```{r}
k_df <- read.csv("/Users/kluzer01/Documents/milo_analysis_2022/Figure_2/k_df3.csv")
k_df2 <- pivot_longer(data.frame(k_df), cols = c(reduced_dimk, graphk), names_to = "method")
k_df2 <- na.omit(k_df2)
ggplot(k_df2, aes(x=(k_val), y=value/60, group_by = method, fill = method)) +
    theme_bw(base_size=16) +
    geom_point(shape=21, size=3) +
    scale_fill_manual(values = c("#8da0cb", "#fc8d62"), name="Sampling scheme", labels = c("Graph", "Reduced dim")) +
    labs(x="k", y="Elapsed time (min)") +
    theme(legend.position = "none", axis.text.x=element_text(colour="black"), axis.text.y=element_text(colour="black")) +
    geom_text(aes(label=round(value/60, 1)),hjust=-0.5,vjust=0.5) +
    scale_y_continuous(limits = c(0,65)) +
    scale_x_continuous(limits = c(0, 100), labels = c("10", "30", "50", "70", "90"), breaks = c(10, 30, 50, 70, 90))

ggsave("fig2c.tiff", path = "/Users/kluzer01/Desktop/figures/", height = 4, width = 4.5, device='tiff', dpi=700)
```

Finally, we import and plot Milo+ speed on a million cell dataset, with 3 different values of k.
To generate this data, run 'benchmark_million.R' using 'benchmark_speed.sh' (change file name).

```{r}
time_df <- read_csv("/Users/kluzer01/Documents/milo_analysis_2022/Figure_2/time_millioncells.csv")
ggplot(time_df, aes(x=as.factor(k_val), y=graph/60)) +
    theme_bw(base_size=16) +
    geom_point(shape=21, size=3, alpha=1, fill = "#8da0cb") +
    labs(x="k", y="Elapsed time (min)") +
    theme(legend.position = "none", axis.text.x=element_text(colour="black"), axis.text.y=element_text(colour="black")) +
    ylim(c(0, 100)) +
    geom_text(aes(label=round(graph/60, 1)),hjust=-0.4,vjust=0.5) 
ggsave("fig2d.tiff", path = "/Users/kluzer01/Desktop/figures/", height = 4, width = 4.5, device='tiff', dpi=700)
```


