
I need to simulate genetically related individuals and their phenotypes to test the GLMM genetic model.

```{r}
library(Matrix)
library(lme4)
library(reshape2)
library(ggplot2)
library(cowplot)
library(glmmTMB)
library(miloR)
library(lme4qtl)
library(MASS)
```
!!! Make sure to source glmm_kin.R !!!

Based on the lme4qtl code, we can create a simulation with known parameter values. 
We can choose the kin dispersion value, and the value of the slope. 

```{r}
simulateFamily <- function(N_ind, size_fam, kin_disp, slope, int) {
    N_ind <- N_ind
    size_fam <- size_fam
    num_fam <- N_ind / size_fam
    
    ### kinship matrix for sib-pairs, we only consider unrelated groups of siblings
    ### specifically, num_fam (eg. 100) families of size_fam (eg. 5) siblings
    kin_fam <- matrix(0.5, size_fam, size_fam) 
    diag(kin_fam) <- 1  
    kin <- kronecker(diag(num_fam), kin_fam)
    	
    ### simulate data: y ~ 0.1*pred1 + 0.1*rand_kin
    # set.seed(1)
    pred1 <- rnorm(N_ind)

    # set.seed(1)
    G <- kin_disp * kin
    b2 <- mvrnorm(n=1, mu=rep(0, N_ind), Sigma=G)
    
    int <- int
    y <- int + (diag(1, N_ind) %*% b2) + slope * pred1 
    
    ### make dataframe
    simdata <- data.frame(id = seq(1, N_ind), fam = rep(seq(1, num_fam), each = size_fam),
      y = y, pred1 = pred1)
    
    rownames(kin) <- seq(1, N_ind)
    colnames(kin) <- seq(1, N_ind)
    
    return(list(kin, simdata))
}

FamSim <- simulateFamily(N_ind = 500, size_fam = 5, kin_disp = 0.8, slope = 0.2, int = 2)

#double check that the simulated data is as expected
#you should be able to predict the slope, intercept (int) and kin_disp values

#slope, int and kin_disp can be predicted using relmatLmer and the kinship matrix stored in FamSim[[1]]
relmat <- relmatLmer(y ~ pred1 + (1|id), FamSim[[2]], relmat = list(id = FamSim[[1]]))
summary(relmat)
```

When using the relmat function to specify the family relationship matrix/additive relationship matrix (A), we get a pretty good estimate of kin_disp.

The next step is to reproduce this for NB distributed count data.
We want to be able to include a conventional random effect, a kinship matrix and an intercept and slope.

Required functions:
```{r}
initializeFullZsim <- function(Z, cluster_levels, stand.cols=FALSE){
    # construct the full Z with all random effect levels
    n.cols <- ncol(Z)
    col.classes <- apply(Z, 2, class)
    i.z.list <- list()
    for(i in seq_len(n.cols)){
        i.class <- col.classes[i]
        if(i.class %in% c("factor")){ # treat as factors
            i.levels <- levels(Z[, i, drop=FALSE])
            i.levels <- as.factor(paste(sort(as.integer(i.levels))))
            i.z <- sapply(i.levels, FUN=function(X) (Z[, i] == X) + 0, simplify=TRUE)
        } else if(i.class %in% c("character")){
            i.levels <- unique(Z[, i, drop=FALSE])
            i.levels <- as.factor(paste(sort(as.integer(i.levels))))
            i.z <- sapply(i.levels, FUN=function(X) (Z[, i] == X) + 0, simplify=TRUE)
        } else if(i.class %in% c("numeric")){ # split into unique levels if integer levels
            i.mod <- all(Z[, i, drop=FALSE] %% 1 == 0)
            if(isTRUE(i.mod)){
                i.levels <- unique(Z[, i])
                i.levels <- as.factor(paste(sort(as.integer(i.levels))))
                i.z <- sapply(i.levels, FUN=function(X) (Z[, i] == X) + 0, simplify=TRUE)
            } else{
                i.z <- Z[, i, drop=FALSE] # if float then treat as continuous
            }
        } else if(i.class %in% c("integer")){
            i.levels <- (unique(Z[, i]))
            i.levels <- as.factor(paste(sort(as.integer(i.levels))))
            i.z <- sapply(i.levels, FUN=function(X) (Z[, i] == X) + 0, simplify=TRUE)
        }
        colnames(i.z) <- cluster_levels[[colnames(Z)[i]]]
        
        # to standardise or not?
        if(isTRUE(stand.cols)){
            q <- ncol(i.z)
            i.ident <- diag(1L, nrow=nrow(i.z), ncol=nrow(i.z))
            i.star <- i.z - ((i.ident %*% i.z)/q)
            i.z <- i.star
        }
        
        i.z.list[[colnames(Z)[i]]] <- i.z
    }
    full.Z <- do.call(cbind, i.z.list)
    return(full.Z)
}
```

I wanted the simulation to be able to:
1. Simulate "typical" random effects (RE1)
2. Simulate random effect related to kinship matrix
3. Perform both these simulations with normal disitrbution 
4. Perform both of these simulations with negative binomial distribution

```{r}
SimulateFamData <- function(N, size_fam, re.sigmas,
                            dispersion, grand.mean, n.re, kin_disp,
                            re.levels, int, slope){
    
    set.seed(2)
    ### kinship matrix for sib-pairs, we only consider unrelated groups of siblings
    ### specifically, num_fam (eg. 100) families of size_fam (eg. 5) siblings
    N <- N
    size_fam <- size_fam
    num_fam <- N / size_fam
    kin_fam <- matrix(0.5, size_fam, size_fam)
    diag(kin_fam) <- 1
    kin <- kronecker(diag(num_fam), kin_fam)
    rownames(kin) <- seq(1, N)
    colnames(kin) <- seq(1, N)
    id <- seq(1, N)
    
    #random u is multivariate normally distributed
    #G = sigma^2 * A, where A is kinship matrix
    #kin_disp defines sigma^2
    G <- kin_disp * kin
    u <- mvrnorm(n=1, mu=rep(0, N), Sigma=G)
    Z.kin <- matrix(0, ncol = N, nrow = N)
    diag(Z.kin) <- 1
    
    if(length(re.levels) != n.re){
        stop("List entries need to match number of input random effects")
    }
    
    #X will just be continuous here 
    pred1 <- rnorm(N)

    #create the design matrix Z for random effects
    #had to rewrite a little to make sure that with very large numbers of random
    #effects, you get at least one individual in each level
    Z <- matrix(0L, ncol=n.re, nrow=N)
    Z[,1] <- sample(c(seq_len(re.levels[[1]]), sample(seq_len(re.levels[[1]]), N-length(seq_len(re.levels[[1]])), replace=TRUE)))
    colnames(Z) <- names(re.levels)

    # construct the full Z
    random.levels <- sapply(seq_len(length(re.levels)), FUN=function(RX) {
        rx.name <- names(re.levels)[RX]
        paste(rx.name, seq_len(re.levels[[rx.name]]), sep="_")
        }, simplify=FALSE)
    names(random.levels) <- names(re.levels)
    full.Z <- initializeFullZsim(Z, random.levels)

    # get a combination over random effects
    # and sample each level from the same ~Normal(0, sigma)
    # note that the variance would be G if we also had random slopes
    re.thetas <- list()
    for(i in seq_len(length(re.levels))){
        i.re <- names(random.levels[i])
        i.levels <- length(random.levels[[i.re]])
        i.re.means <- rnorm(n=i.levels, 0, sd=sqrt(re.sigmas[[i.re]])) # sample a random effect value
        i.re.list <- sapply(seq_len(i.levels), FUN=function(X) i.re.means[X])
        names(i.re.list) <- random.levels[[i.re]]
        re.thetas[[i.re]] <- i.re.list
    }
    
    # construct the y.means equation, depending on desired distribution and FE/RE
    #use this for a normal distribution 
    #y.means <- int + (slope*pred1) + (full.Z %*% unlist(re.thetas)) + (Z.kin %*% u) + rnorm(id, 0, 0.0001)
    #y.counts <- rep(NA, N)
    
    #use this for a negative binomial distribution
    y.means <- exp(int + (slope*pred1) + (full.Z %*% unlist(re.thetas)) + (Z.kin %*% u))
    #y.means <- exp(int + (slope*pred1) + (Z.kin %*% u))
    y.counts <- rnbinom(N, mu = y.means, size = dispersion)
    
    sim.data <- data.frame("Mean"=y.means, "Mean.Count"=y.counts, "id"=id, 
                           "fam" = rep(seq(1, num_fam), each = size_fam), "pred1" = pred1)
    sim.data <- do.call(cbind.data.frame, list(sim.data, Z))
    return(list(sim.data, kin))
}

simulated_genetic_data <- SimulateFamData(N=500, size_fam=50, re.sigmas=list("RE1"=0.2),
                            dispersion=0.5, grand.mean=2, n.re=1, kin_disp=0.3, int = 2,
                            slope=0.4, re.levels=list("RE1"=10))

# check with TMB that the RE1 variance is as expected (0.2 in this case)
#will probably be a bit more because we also introduce variance with the kinship

glmm.tmb <- glmmTMB(Mean.Count ~ pred1 + (1|RE1), simulated_genetic_data[[1]], family=nbinom2(link="log"), REML = TRUE)
summary(glmm.tmb)
```

Here I redid the simulation to create a custom covariance matrix for between families. 
It was done manually, might be better to rewrite this. 
```{r}
SimulateFamData2 <- function(N, size_fam, re.sigmas,
                            dispersion, grand.mean, n.re, kin_disp,
                            re.levels, int, slope){
    
    set.seed(2)
    ### kinship matrix for sib-pairs, we only consider unrelated groups of siblings
    ### specifically, num_fam (eg. 100) families of size_fam (eg. 5) siblings
    N <- N
    size_fam <- size_fam
    num_fam <- N / size_fam
    kin_family <- matrix(0, num_fam, num_fam)
    diag(kin_family) <- 1
    kin_family[1,2] <- kin_family[2,1] <- 0.5
    kin_family[5,6] <- kin_family[6,5] <- 0.5
    kin_family[4,3] <- kin_family[3,4] <- 0.5
    kin_family[7,8] <- kin_family[8,7] <- 0.5
    kin_family[9,10] <- kin_family[10,9] <- 0.5

    id <- seq(1, N)
    
    G <- kin_disp * kin_family
    u <- mvrnorm(n=1, mu=rep(0, num_fam), Sigma=G)
    fam.id <- matrix(rep(seq(1, num_fam), each = size_fam), ncol=1)
    colnames(fam.id) <- "fam"
    re.levels.fam <- list("fam"=10)
    random.levels.fam <- sapply(seq_len(length(re.levels.fam)), FUN=function(RX) {
        rx.name <- names(re.levels.fam)[RX]
        paste(rx.name, seq_len(re.levels.fam[[rx.name]]), sep="_")
        }, simplify=FALSE)
    names(re.levels.fam) <- names(re.levels.fam)
    Z.kin <- initializeFullZsim(fam.id, random.levels.fam)
    
    if(length(re.levels) != n.re){
        stop("List entries need to match number of input random effects")
    }
    
    #X will just be continuous here 
    pred1 <- rnorm(N)

    #create the design matrix Z for random effects
    Z <- matrix(0L, ncol=n.re, nrow=N)
    Z[,1] <- sample(c(seq_len(re.levels[[1]]), sample(seq_len(re.levels[[1]]), N-length(seq_len(re.levels[[1]])), replace=TRUE)))

    colnames(Z) <- names(re.levels)

    # construct the full Z
    random.levels <- sapply(seq_len(length(re.levels)), FUN=function(RX) {
        rx.name <- names(re.levels)[RX]
        paste(rx.name, seq_len(re.levels[[rx.name]]), sep="_")
        }, simplify=FALSE)
    names(random.levels) <- names(re.levels)
    full.Z <- initializeFullZsim(Z, random.levels)

    # get a combination over random effects
    # and sample each level from the same ~Normal(0, sigma)
    # note that the variance would be G if we also had random slopes
    re.thetas <- list()
    for(i in seq_len(length(re.levels))){
        i.re <- names(random.levels[i])
        i.levels <- length(random.levels[[i.re]])
        i.re.means <- rnorm(n=i.levels, 0, sd=sqrt(re.sigmas[[i.re]])) # sample a random effect value
        i.re.list <- sapply(seq_len(i.levels), FUN=function(X) i.re.means[X])
        names(i.re.list) <- random.levels[[i.re]]
        re.thetas[[i.re]] <- i.re.list
    }
    
    # construct the y.means equation, depending on desired distribution and FE/RE
    #y.means <- int + (slope*pred1) + (full.Z %*% unlist(re.thetas)) + (Z.kin %*% u) + rnorm(id, 0, 0.0001)
    #y.counts <- rep(NA, N)
    #y.means <- exp(int + (slope*pred1) + (full.Z %*% unlist(re.thetas)) + (Z.kin %*% u))
    y.means <- exp(int + (slope*pred1) + (Z.kin %*% u))
    y.counts <- rnbinom(N, mu = y.means, size = dispersion)
    
    sim.data <- data.frame("Mean"=y.means, "Mean.Count"=y.counts, "id"=id, 
                           "fam" = rep(seq(1, num_fam), each = size_fam), "pred1" = pred1)
    sim.data <- do.call(cbind.data.frame, list(sim.data, Z))
    return(list(sim.data, kin_family))
}

simulated_family_data <- SimulateFamData2(N=500, size_fam=50, re.sigmas=list("RE1"=0.2),
                            dispersion=0.5, grand.mean=2, n.re=1, kin_disp=0.3, int = 2,
                            slope=0.4, re.levels=list("RE1"=10))

#in this case I have set the RE1 to be 0 (ie I have not included it in the model),
#so this should predict an RE1 close to 0
glmm.tmb <- glmmTMB(Mean.Count ~ pred1 + (1|RE1), simulated_family_data[[1]], family=nbinom2(link="log"), REML = TRUE)
summary(glmm.tmb)
```

And now, for the main bit... 
We want to incorporate the kinship matrix in our NB-glmm model 

The changes to the NB-glmm model required to do this are listed below. 
There are all tagged with "#####here"
1) change the rownames of curr_sigma to genetic, so that sigma gets multiplied by A in initialiseGR
2) rename the variable random levels "genetic", so that sigma gets multiplied by A in initaliseGR
3) add the kinship matrix as a variable in initialiseGR, and compute G = sigma*A 
4) create a function called computeV_partial_kin, which includes the kinship matrix

I show that this works when the number of levels of the kinship matrix is small, using the simulated 
family data. 

```{r}
glmm.tmb <- glmmTMB(Mean.Count ~ pred1 + (1|RE1), simulated_family_data[[1]], family=nbinom2(link="log"), REML = TRUE)
summary(glmm.tmb)

#random.levels <- list("RE1"=unique(as.numeric(as.factor(simulated_family_data[[1]]$RE1))))
random.levels <- NULL
X <- as.matrix(data.frame("Intercept"=rep(1, nrow(simulated_family_data[[1]])), "FE2"=as.numeric(simulated_family_data[[1]]$pred1)))
Z <- as.matrix(data.frame("RE1"=as.numeric(simulated_family_data[[1]]$fam)))
y <- simulated_family_data[[1]]$Mean.Count
dispersion <- 0.5

glmm.control <- glmmControl.defaults()
glmm.control$theta.tol <- 1e-5
glmm.control$max.iter <- 50

model.list <- runGLMMkin(X=X, Z=Z, y=y, random.levels=random.levels, REML = TRUE,
                              glmm.control=glmm.control, dispersion=dispersion, Kin=simulated_family_data[[2]])

model.coef <- c(model.list$FE, model.list$Sigma)
model.coef
```
As expected, the intercept is about 2, the slope is about 0.4, and the genetic variance is about 0.3 (a bit higher - but there is definitely more variance explained than when modeling RE1).

However, when the kinship matrix contains more levels (approaching N), we do not get convergence. This behaviour can also be seen when modeling typical random effects (RE1) with a large number of levels. 

```{r}
glmm.tmb <- glmmTMB(Mean.Count ~ pred1 + (1|RE1), simulated_genetic_data[[1]], family=nbinom2(link="log"), REML = TRUE)
summary(glmm.tmb)

#random.levels <- list("RE1"=unique(as.numeric(as.factor(simulated_genetic_data[[1]]$id))))
random.levels <- NULL
X <- as.matrix(data.frame("Intercept"=rep(1, nrow(simulated_genetic_data[[1]])), "FE2"=as.numeric(simulated_genetic_data[[1]]$pred1)))
Z <- as.matrix(data.frame("RE1"=as.numeric(simulated_genetic_data[[1]]$id)))
y <- simulated_genetic_data[[1]]$Mean.Count
dispersion <- 0.5

glmm.control <- glmmControl.defaults()
glmm.control$theta.tol <- 1e-5
glmm.control$max.iter <- 50

model.list <- runGLMMkin(X=X, Z=Z, y=y, random.levels=random.levels, REML = TRUE,
                              glmm.control=glmm.control, dispersion=dispersion, Kin=simulated_genetic_data[[2]])

model.coef <- c(model.list$FE, model.list$Sigma)
model.coef
```

```{r}
conv.list <- model.list$conv.list
diff.df <- melt(do.call(cbind,
                        lapply(conv.list, function(X){
                            if(!is.null(ncol(X$theta_diff))){
                                x <- X$theta_diff[, 1]
                                names(x) <- rownames(X$Theta)
                            } else{
                               x <- X$theta_diff
                               names(x) <- c(colnames(X), paste(colnames(Z), c(1:length(levels(as.factor(simulated_genetic_data[[1]]$id)))), sep="_"))
                            }
                            x})))
diff.df$Var2 <- rep(1:(length((model.list$conv.list))), each = 502)

diff.df2 <- melt(do.call(cbind,
                        lapply(conv.list, function(X){
                            if(!is.null(ncol(X$sigma_diff))){
                                x <- X$sigma_diff[, 1]
                                names(x) <- rownames(X$Sigma)
                            } else{
                               x <- X$sigma_diff
                               names(x) <- c(colnames(X), paste(colnames(Z), c(1:length(levels(as.factor(simulated_genetic_data[[1]]$id)))), sep="_"))
                            }
                            x})))
diff.df2$Var2 <- rep(1:(length((model.list$conv.list))), each = 1)

diff.df3 <- melt(do.call(cbind,
                        lapply(conv.list, function(X){
                            if(!is.null(ncol(X$loglihood))){
                                x <- X$loglihood[, 1]}})))
diff.df3$Var2 <- rep(1:(length((model.list$conv.list))), each = 1)

theta.conv <- 1e-5

diff.df <- dplyr::filter(diff.df, Var1 == c("Intercept", "FE2"))
ggplot(diff.df, aes(x=Var2, y=abs(value), colour=as.factor(Var1))) +
    geom_hline(yintercept=theta.conv, lty=2, col='red') +
    geom_line() +
    theme_cowplot() +
    labs(x="Iteration", y=expression(theta[0] - theta["t"]), colour = "Variable") +
    facet_wrap(~Var1, scales="free_y", ncol = 6, labeller = as_labeller(labels)) +
    theme(legend.position = "none") +
    NULL
ggplot(diff.df2, aes(x=Var2, y=abs(value), colour=as.factor(Var1))) +
    geom_hline(yintercept=theta.conv, lty=2, col='red') +
    geom_line() +
    theme_cowplot() +
    labs(x="Iteration", y=expression(sigma[0] - sigma["t"]), colour = "Variable") +
    facet_wrap(~Var1, scales="free_y", ncol = 6, labeller = as_labeller(labels)) +
    theme(legend.position = "none") +
    NULL
ggplot(diff.df3, aes(x=Var2, y=abs(value), colour=as.factor(Var1))) +
    geom_line() +
    theme_cowplot() +
    labs(x="Iteration", y="log-likelihood", colour = "Variable") +
    facet_wrap(~Var1, scales="free_y", ncol = 6, labeller = as_labeller(labels)) +
    theme(legend.position = "none") +
    NULL
```

As we can see, we get a weird bouncing behaviour. 

Using a regular random effect, RE1, I tried to determine the limits of this behaviour. How many random levels can the model withstand before it becomes unstable?

```{r}
j <- 1
sequence <- seq(10, 500, 10)
iters <- rep(NA, length(sequence))
conv <- rep(NA, length(sequence))
sigma_val <- list()

for (i in sequence) {
  print(i)
  simulated_genetic_data <- SimulateFamData(N=500, size_fam=10, re.sigmas=list("RE1"=0.2),
                              dispersion=0.5, grand.mean=2, n.re=1, kin_disp=0.1, int = 2,
                              slope=0.4, re.levels=list("RE1"=i))
  
  random.levels <- list("RE1"=unique(as.numeric(as.factor(simulated_genetic_data[[1]]$RE1))))
  #random.levels <- NULL
  X <- as.matrix(data.frame("Intercept"=rep(1, nrow(simulated_genetic_data[[1]])), "FE2"=as.numeric(simulated_genetic_data[[1]]$pred1)))
  Z <- as.matrix(data.frame("RE1"=as.numeric(simulated_genetic_data[[1]]$RE1)))
  y <- simulated_genetic_data[[1]]$Mean.Count
  dispersion <- 0.5
  kin <- list(id = simulated_genetic_data[[2]])
  
  glmm.control <- glmmControl.defaults()
  glmm.control$theta.tol <- 1e-4
  glmm.control$max.iter <- 70
  
  model.list <- runGLMMkin(X=X, Z=Z, y=y, random.levels=random.levels, REML = TRUE,
                                glmm.control=glmm.control, dispersion=dispersion, Kin=simulated_genetic_data[[2]])
  conv.list <- model.list$conv.list
  sigma_val[[j]] <- lapply(conv.list, `[[`, "Sigma")
  iters[j] <- model.list$Iters
  conv[j] <- model.list$converged
  j <- j + 1
}

```

Finally, I looked at various methods to induce stability. I tried two main approaches to do this.

Firstly, by applying the conjugate gradient method (CGM) and the CGM + pre-conditioner matrix M to solve
the linear system of equations in the mixed model equations (MMEs). I tried:
Cx=r
- Just CGM
- CGM + preconditioner M, where M is a diagonal matrix (diag(C))
- CGM + preconditioner M, where M is an incomplete cholesky factorisation of C

I also tried to diagonalize the MMEs with the two following methods:
- Z* = ZL, u* = Lu (cholesky transformed)
- G = VDV’, D-1 instead (eigenvalue decomposition)
More details can be found in this publication: https://onlinelibrary.wiley.com/doi/epdf/10.1046/j.1439-0388.2002.00327.x

My implementations of these methods can be found in glmm_kin.R:
- solve_equations_kin (bottom, commented out)
- solve_equations_kin2 cholesky
- solve_equations_kin3 eigenvalue decomposition
