---
title: "Benchmarking DA methods"
output: html_notebook
---

```{r}
library(tidyverse)
library(MouseGastrulationData)
library(SingleCellExperiment)
library(scater)
library(scran)
library(bit64)
library(glue)
library(patchwork)
library(ggpubr)

source('/Users/kluzer01/Documents/milo_analysis_2022/Figure_1/benchmark/benchmark_utils.R')
```

Some plotting utils

```{r}
## color palette for methods
method_names <- c("milo_red", "milo_tri")
method_colors <- RColorBrewer::brewer.pal(7, name="Spectral")[c(5,6)]
method_colors <- c("#fc8d62", "#8da0cb")
method_colors <- setNames(method_colors, method_names)
method_labels <- c("Reduced_dim", "Graph")
method_labels <- setNames(method_labels, method_names)

scale_color_methods <- function(){ scale_color_manual(values = method_colors, labels=method_labels) }
scale_fill_methods <- function(){ scale_fill_manual(values = method_colors, labels=method_labels) }

remove_x_axis <- function(){
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.title.x = element_blank())  
}

remove_y_axis <- function(){
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.title.y = element_blank())  
}
```

```{r}
enr_labeller2 <- function(value){
  fc <- round(prob2FC(as.numeric(value)), 2)
  enr_name <- paste("logFC = ", fc)
  return(c(value=enr_name))
}

pl_labeller <- labeller(method=as_labeller(setNames(method_labels, method_names)),
     enr=enr_labeller2)
```

In this notebook I will visualize the results from benchmarking Milo reduced_dim against graph.

## Benchmarking workflow

All the main functions and scripts are implemented in `benchmark`. To run the analysis, several scripts need to be called:

- Condition labels are simulated by running `make_bm_data.R` (wrapped in `submit_make_bm_data.sh`)
- Methods implemented in R are run with `run_DA_R.r` (wrapped in `submit_benchmark.sh`)
- Finally, the results are collated with `build_outcome.R`

This document will outline all the necessary steps.

## Define input datasets

These are going to be used as underlying KNN graphs with different topologies for benchmarking.

### Mouse gastrulation data

Select samples from late time points (even number of replicates)

```{r}
AtlasSampleMetadata %>%
  arrange(stage) %>%
  mutate(sample=factor(sample, levels=unique(sample))) %>%
  ggplot(aes(stage, sample)) +
  geom_point()

late_samples <- AtlasSampleMetadata %>%
  filter(stage %in% c("E7.75", "E8.0", "E8.25", "E8.5")) %>%
  pull(sample)
```

```{r}
embryo_sce <- EmbryoAtlasData(type="processed", samples = late_samples)
```

```{r}
data.frame(colData(embryo_sce)) %>%
  ggplot(aes(as.factor(sample), as.factor(pool))) + geom_point()
```

Preprocessing

```{r}
logcounts(embryo_sce) <- log1p(counts(embryo_sce))
## Exclude zero counts genes
keep.rows <- rowSums(logcounts(embryo_sce)) != 0
embryo_sce <- embryo_sce[keep.rows, ]

dec <- modelGeneVar(embryo_sce)
hvgs <- getTopHVGs(dec, n=5000)

## Drop cells with NAs in corrected pca (low quality)
embryo_sce <- embryo_sce[,apply(reducedDim(embryo_sce, "pca.corrected"), 1, function(x) all(!is.na(x)))]

## Run UMAP (the data has been subsetted)
embryo_sce <- scater::runUMAP(embryo_sce, dimred="pca.corrected")
```

```{r, fig.height=6, fig.width=10}
plotReducedDim(embryo_sce, "UMAP", colour_by="celltype", text_by="celltype", point_size=0.1) +
  scale_color_manual(values=EmbryoCelltypeColours) +
  guides(color=guide_legend(override.aes = list(size=2)))
```

Save object for benchmark

```{r}
saveRDS(embryo_sce, "/Users/kluzer01/Documents/milo_analysis_2022/Figure_1/benchmark/embryo_data_bm.RDS")
```

```{r}
embryo_sce <- readRDS("/Users/kluzer01/Documents/milo_analysis_2022/Figure_1/benchmark/embryo_data_bm.RDS")
```

Save list of celltype names and sample by sizes.

```{r}
for (i in 1:3) {
  pop_size_df <- as.data.frame(table(embryo_sce$celltype)) %>%
  dplyr::rename(pop=Var1, pop_size=Freq) %>%
  mutate(size_bin = cut(pop_size, breaks = 10)) %>%
  group_by(size_bin) %>%
  sample_n(size=1) 
  pop_size_df$pop <- gsub(" ", "_", pop_size_df$pop)
  pop_size_df <- pop_size_df[-which(pop_size_df$pop == "Forebrain/Midbrain/Hindbrain"),]
  write(as.character(pop_size_df$pop), glue("/Users/kluzer01/Documents/milo_analysis_2022/Figure_1/benchmark/data/pop_sample_{i}.txt"))
}
```


## At this point, you have finished creating the population files. The analysis continues with the following steps:
1. Create a conda environment on cluster and name it milo_bm.
2. Copy the milo_analysis_2022 repository onto the cluster
3. Run 'submit_make_bm_data.sh'. Make sure you activate read/write/execute permissions! This creates DA simulations for the different populations of cells. This step should only take a few minutes.
4. Run 'submit_benchmark.sh'. This runs the two methods we are trying to compare (reduced_dim and graph) on the simulations. This step will take a few hours (~4 hours).
5. Copy all /data/output_benchmark/benchmark_* files to a local directory.

## Pick threshold for cells in true DA regions based on p(C1) distribution

Read full outcomes and true probabilities

```{r}
read4_prob_hist <- function(data_id){
  outdir <- glue("/Users/kluzer01/Documents/milo_analysis_2022/Figure_1/benchmark/data/output_benchmark/")
  res_files <- list.files(outdir, pattern=glue("embryo.+batchEffect0.DAresults"))
  res_files_full <- list.files(outdir, pattern=glue("embryo.+batchEffect0.DAresults"), full.names = TRUE)
  #res_files_full_nomeld <- str_subset(res_files_full, "meld|cydar_batch|louvain_batch|milo_batch", negate = TRUE)
  
  res_ls <- lapply(seq_along(res_files_full), function(i){ 
    print(paste("Outcome no. ", i))
    benchmark_df <- read_csv(res_files_full[i]) 
    ## Check if conditions were swapped in test
    pred_cor <- benchmark_df %>%
      mutate(pred=factor(pred, levels=c("NegLFC", "NotDA", "PosLFC"), ordered = TRUE)) %>%
      mutate(pred=as.numeric(pred)) %>%
      summarise(cor(pred, true_prob))
    ## Swap outcomes if so
    if (!is.na(pred_cor)) {
      if (pred_cor < - 0.1) {
        benchmark_df <- mutate(benchmark_df, pred = ifelse(pred=="NegLFC", "PosLFC", ifelse(pred=="PosLFC", "NegLFC", "NotDA")))
      }
    }
    benchmark_df %>%
      mutate(id=i)
    })
  
  res_df <- res_ls  %>%
    purrr::reduce(bind_rows) 
  # %>%
  #   select(true_prob, pred, method) 
  
  return(res_df)  
}

embryo_res_df <- read4_prob_hist("embryo")
```

Plot distribution of P(C1) in cells identified as DA by different methods

```{r, fig.height=8, fig.width=18}
calculate_quantile_thresh <- function(res_df){
  return(quantile(1-res_df[res_df$method=="milo_red",][["true_prob"]], probs = seq(0,1, by=0.25))["75%"])
}

plot_p_hist <- function(res_df){
  ## Calculate threshold
  thresh <- calculate_quantile_thresh(res_df)
  
  ## Plot histogram of p(C1) for predicted DA cells
  p <- res_df %>%
    dplyr::filter(pred!="NotDA") %>%
    # mutate(color=ifelse(true_prob==0.5, "0.5","0")) %>%
    ggplot(aes(1 - true_prob)) +
    # geom_density(aes(color=pred)) +
    facet_grid(method~pred, scales="free_y", labeller=pl_labeller) +
    geom_histogram(bins=50, aes(fill=pred)) +
    xlab("P(C1)") + ylab("# DA cells") +
    theme_bw(base_size=16) +
    scale_fill_brewer(palette="Set1", name="Predicted effect") +
    geom_vline(xintercept = thresh, linetype=2) 
  p
}

# plot_p_hist(cluster_res_df) + ggtitle("clusters") 
plot_p_hist(embryo_res_df) + ggtitle("embryo") +
  plot_layout(guides="collect")
```

```{r, fig.height=5, fig.width=16}
plot_thresh_hist <- function(linear_res_df){
  quants <- quantile(1-linear_res_df$true_prob, seq(0, 1, 0.05))
  keep_quants <- c("25%", "50%",  "90%", "95%")
  linear_res_df %>%
    filter(method=="milo") %>%
      ggplot(aes(1 - true_prob)) +
      geom_histogram(bins=50) +
      xlab("P(C1)") + ylab("# cells") +
      theme_bw(base_size=16) +
      scale_fill_brewer(palette="Set1", name="Predicted effect") +
      geom_vline(xintercept = quants[keep_quants], linetype=2) +
    coord_cartesian(clip = 'off') +
    annotate('text', x=quants[keep_quants], y=Inf, label=(keep_quants), hjust=0, vjust=0, angle=60) +
    geom_vline(xintercept = quants["75%"], linetype=2, color="red") +
    annotate('text', x=quants["75%"], y=Inf, label="75%", hjust=0, vjust=0, angle=60, color="red") +
    annotate('text', x=Inf, y=Inf, label=paste('t =', round(quants["75%"], 3)), hjust=1.1, vjust=1.5, angle=0, color='red', size=5) +
    theme(plot.margin = unit(c(1,0.5,1,0.5), "cm"), title = element_text(vjust = 1)) 
}

plot_thresh_hist(embryo_res_df) + ggtitle("Mouse Gastrulation", subtitle="")
```

Use 75% quantile of P(C1) in all cells to define threshold for true DA

```{r}
embryo_thresh <- calculate_quantile_thresh(embryo_res_df)
```

Some utils

```{r}
.outcome_by_prob <- function(benchmark_df, da_upper){
  DA_thresh <- 1 - da_upper
  benchmark_df <- benchmark_df %>%
    mutate(true = ifelse(benchmark_df$true_prob < DA_thresh, "NegLFC", ifelse(benchmark_df$true_prob > 1-DA_thresh, "PosLFC", "NotDA"))) 
  if (benchmark_df$method[1]=="meld") {
    benchmark_df <- mutate(benchmark_df, pred = ifelse(Condition2 < DA_thresh, "NegLFC", "NotDA")) 
  }
  ## Check if conditions were swapped in test
  pred_cor <- benchmark_df %>%
    group_by(pred) %>%
    summarise(m=mean(true_prob)) %>%
    ungroup() %>%
    mutate(pred=factor(pred, levels=c("NegLFC", "NotDA", "PosLFC"), ordered = TRUE)) %>%
    mutate(pred=as.numeric(pred)) %>%
    summarise(c=cor(pred, m)) %>%
    pull(c)
  ## Swap outcomes if so
  if (!is.na(pred_cor)) {
    if (pred_cor < 0) {
      benchmark_df <- mutate(benchmark_df, pred = ifelse(pred=="NegLFC", "PosLFC", ifelse(pred=="PosLFC", "NegLFC", "NotDA")))
    }
  }
  calculate_outcome(benchmark_df)
}

plot_outcome_umap <- function(sce, method, pop, enr, seed, batchEffect, true=FALSE, data_id="embryo", 
                              red_dim="umap_batch",
                              rasterize=FALSE){
  coldata_file <- list.files(indir, pattern=paste0(".+",data_id,"_pop_",  pop,"_enr",enr,"_seed", seed, ".coldata.csv"),
                           full.names = TRUE)
  colData(sce) <- read_csv(coldata_file) %>% column_to_rownames() %>% DataFrame()
  
  res_file <- list.files(outdir, 
                         pattern= paste0(".+",data_id,"_pop_", pop,"_enr",enr,"_seed", seed, '.+batchEffect',batchEffect,".DAresults.",method,".csv"),
                         full.names = TRUE)
  if (method=="meld") {
    sce$predicted <- ifelse(read_csv(res_file)[["Condition2"]] < 0.3, "NegLFC", 'NotDA')  
  } else if (method=="daseq") {
    sce$predicted <- ifelse(read_csv(res_file)[["pred"]] == "NegLFC", 'PosLFC', ifelse(read_csv(res_file)[["pred"]] == "PosLFC", "NegLFC", "NotDA"))
  } else {
    sce$predicted <- read_csv(res_file)[["pred"]]
  }
  if (true) {
    pl <- data.frame(reducedDim(sce, red_dim)) %>%
    dplyr::rename_all( ~ c("UMAP1", "UMAP2") ) %>%
    mutate(true=sce$Condition1_prob) %>%
    arrange(true) %>%
    ggplot(aes(UMAP1, UMAP2, color=true)) + 
    geom_point(size=0.1) +
      scale_color_viridis_c(name="C1 probability", limits=c(0.5,0.9)) +
     theme_classic(base_size=18) +
      guides(color=guide_colorbar(title.position="top", title.hjust = 0.5)) +
     theme(axis.ticks = element_blank(),
           axis.text = element_blank(),
           legend.position = "top") +
     xlab("UMAP1") + ylab("UMAP2") 
  } else {
    pl <- data.frame(reducedDim(sce, red_dim)) %>%
    dplyr::rename_all( ~ c("UMAP1", "UMAP2") ) %>%
    dplyr::mutate(method=method) %>%
    dplyr::mutate(predicted=factor(sce$predicted, levels=c("NegLFC", "PosLFC", 'NotDA'))) %>%
    dplyr::arrange(- predicted) %>%
    ggplot(aes(UMAP1, UMAP2, color=predicted)) + 
    geom_point(size=0.1) +
    geom_point(data= . %>% filter(predicted!="NotDA"), size=0.2) +
     scale_color_manual(values=c(NegLFC="red", NotDA="grey", PosLFC="blue"), 
                        labels=c(NegLFC="Enriched in C1", NotDA="No DA", PosLFC="Depleted in C1"), name='') +
     guides(color = guide_legend(override.aes = list(size=1))) +
     theme_classic(base_size=18) +
     theme(legend.position=c(.15,.2), axis.ticks = element_blank(),
           axis.text = element_blank()) +
      facet_wrap(method~.)  
  }
  if (rasterize) {
    pl <- pl + rasterize(geom_point(size=0.1), dpi = 300)
  }
  pl
}

## Convert condition probability to LFC
prob2FC <- function(prob){
  log((prob)/(1 - prob))
}
```

Next step:

1. Run 'build_outcome.R'. Use embryo_thresh value to set threshold in build_outcome.R.
2. Copy output file 'outcome_embryo_batchEffect0.csv' to a local directory.

## Embryo data

### W/o batch effect

Read outcome for embryo (built with wrapper `build_outcome.R`).

```{r}
outcome_df <- read_csv("/Users/kluzer01/Documents/milo_analysis_2022/Figure_1/benchmark/data/output_benchmark/outcome_embryo_batchEffect0.csv")
```

```{r}
# outcome_df2 <- outcome_df %>%
#   filter(DA_thresh==embryo_thresh) %>%
#   filter(pop!="Mixed_mesoderm")
```


```{r, fig.width=12, fig.height=5}
# enr_labeller2 <- function(variable, value, split_lines=FALSE){
#   fc <- round(prob2FC(as.numeric(value)), 2)
#   enr_name <- paste("logFC = ", fc)
#   return(enr_name)
# }
pl_df <- outcome_df %>% 
   filter(batchEffect==0 & str_detect(method, "_batch", negate = TRUE)) %>% 
   mutate(FDR = FP/(TP+FP)) %>% 
   mutate(TPR=ifelse(is.nan(TPR), 0, TPR), 
          FPR=ifelse(is.nan(FPR), 0, FPR), 
          FDR=ifelse(is.nan(FDR), 0, FDR), 
          Precision=ifelse(is.nan(Precision), 0, Precision)) %>% 
   filter(batchEffect==0)  %>% 
   mutate(enr=as.character(enr))  

p_top <- pl_df %>%
  filter(batchEffect=="0") %>%
  ggplot(aes(as.factor(round(prob2FC(as.numeric(enr)),1)), TPR, color=method)) +
  geom_boxplot(outlier.alpha = 0, alpha =1) +
  ggbeeswarm::geom_quasirandom(alpha=1, position="stack", size=0.5) +
  # geom_hline(yintercept = 0.1, linetype=2) +
  theme_bw(base_size = 12) +
  scale_color_manual(values = method_colors, labels=setNames(method_labels, method_names)) +
  scale_fill_manual(values = method_colors, labels=setNames(method_labels, method_names)) +
  theme(legend.position = "none", axis.text.x=element_text(colour="black"), axis.text.y=element_text(colour="black")) +
  facet_grid(.~method, labeller=pl_labeller) +
  #theme(strip.background =element_rect(fill="grey")) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.title.x = element_blank())

p_bottom <- pl_df %>%
  filter(batchEffect=="0") %>%
  ggplot(aes(as.factor(round(prob2FC(as.numeric(enr)),1)), FDR, color=method)) +
  geom_boxplot(outlier.alpha = 0, alpha =1) +
  ggbeeswarm::geom_quasirandom(alpha=1, position="stack", size=0.5) +
  geom_hline(yintercept = 0.1, linetype=2) +
  theme_bw(base_size = 12) +
  scale_color_manual(values = method_colors, labels=setNames(method_labels, method_names)) +
  scale_fill_manual(values = method_colors, labels=setNames(method_labels, method_names)) +
  theme(legend.position = "none", axis.text.x=element_text(colour="black"), axis.text.y=element_text(colour="black")) +
  facet_grid(.~method, labeller=pl_labeller) +
  xlab('Simulated log fold change') +
  ylim(c(0, 1))

(p_top / p_bottom) + plot_layout(guides="collect") 
ggsave("fig1cc.tiff", path = "/Users/kluzer01/Desktop/figures/", height = 4.5, width = 5, device='tiff', dpi=700)

```


#### Explain variability in performance across different populations

Due to low signal to noise ratio for certain simulations

```{r}
## Read outputs
outdir <- "/Users/kluzer01/Documents/milo_analysis_2022/Figure_1/benchmark/data/output_benchmark"
res_files <- list.files(outdir, pattern=".DAresults")
res_files_full <- list.files(outdir, pattern=".DAresults", full.names = TRUE)

## Make data frame w benchmark parameters
out_meta_df <- data.frame(file_id = str_remove_all(res_files, "benchmark_embryo_pop_|.csv")) %>%
  separate(col = file_id, sep = ".DAresults.", into=c("file_id", "method")) %>%
  separate(col = file_id, sep = "_enr", into=c("pop", "file_id")) %>%
  separate(col = file_id, sep = "_", into=c("enr", "seed", "batchEffect")) %>%
  mutate(seed=str_remove(seed, "seed"), batchEffect=str_remove_all(batchEffect, "batchEffect")) 

embryo_thresh <- 0.5406022
no_batch_files <- str_detect(res_files_full, "batchEffect0.DAresult") & !str_detect(res_files_full, "louvain_batch|cydar_batch|milo_batch")

## Read benchmark results w/o batch effects and selecting just top and bottom population
keep_pops <- c("Erythroid2", "Caudal_neurectoderm")
keep_files <- str_detect(res_files_full, "batchEffect0.DAresult") & 
  !str_detect(res_files_full, "louvain_batch|cydar_batch|milo_batch") &
  str_detect(res_files_full, paste0(keep_pops, collapse = "|"))
suppressWarnings({
  outcome_df_2 <- lapply(seq_along(res_files_full[keep_files]), function(i){ 
    print(paste("Outcome no. ", i))
    benchmark_df <- read_csv(res_files_full[keep_files][i])
    pop_enr <- as.numeric(out_meta_df[keep_files,][i,"enr"])
    benchmark_df <- benchmark_df %>%
      mutate(true = ifelse(benchmark_df$true_prob < 1-embryo_thresh, "NegLFC", ifelse(benchmark_df$true_prob > embryo_thresh, "PosLFC", "NotDA"))) 
    
    ## Check if conditions were swapped in test
    pred_cor <- benchmark_df %>%
      group_by(pred) %>%
      summarise(m=mean(true_prob)) %>%
      ungroup() %>%
      mutate(pred=factor(pred, levels=c("NegLFC", "NotDA", "PosLFC"), ordered = TRUE)) %>%
      mutate(pred=as.numeric(pred)) %>%
      summarise(c=cor(pred, m)) %>%
      pull(c)
    ## Swap outcomes if so
    if (!is.na(pred_cor)) {
      if (pred_cor < 0) {
        benchmark_df <- mutate(benchmark_df, pred = ifelse(pred=="NegLFC", "PosLFC", ifelse(pred=="PosLFC", "NegLFC", "NotDA")))
      }
    }
   
    benchmark_df %>%
      mutate(prob_bins=cut(1-true_prob, breaks = seq(0.3, 1, by=0.05), include.lowest = TRUE, right=FALSE)) %>%
        group_by(prob_bins) %>%
        do(calculate_outcome(.)) %>%
        ungroup() %>%
        dplyr::select(- method) %>%
        bind_cols(out_meta_df[keep_files,][i,]) 
    }) %>%
    purrr::reduce(bind_rows)
})
outcome_df_2
```


```{r, fig.height=7, fig.width=12}
pl_df <- outcome_df_2 %>% 
  mutate(prob_bins= seq(0.3, 1, by=0.05)[as.numeric(outcome_df_2$prob_bins)]) %>%
   filter(batchEffect==0 & str_detect(method, "_batch", negate = TRUE)) %>% 
   mutate(TPR=TP/(TP+FN), FPR=FP/(FP+TN)) %>% 
   mutate(TPR=ifelse(is.nan(TPR), 0, TPR), 
          FPR=ifelse(is.nan(FPR), 0, FPR),
          ) %>% 
   # filter(prob_bins >= 0.55 & batchEffect==0)  %>% 
   mutate(enr=as.character(enr))  %>%
  filter(pop %in% c("Erythroid2", "Caudal_neurectoderm"))

p_top <- pl_df %>%
  filter(batchEffect=="0") %>%
  ggplot(aes(as.factor(prob_bins), TPR, color=pop, shape=pop)) +
  ggbeeswarm::geom_quasirandom(alpha=0.3, position="stack", size=0.5) +
  geom_point(data = . %>% group_by(method, pop, prob_bins) %>% summarise(TPR=mean(TPR)), size=2,
                          position=position_dodge(width=0.5)) +
  theme_bw(base_size = 12) +
 #scale_shape_manual("Population", labels = c("X", "Y")) +
  facet_grid(.~method, labeller=pl_labeller) + 
  theme(axis.text.x = element_text(angle=90, colour="black"), legend.position = "none", axis.text.y=element_text(colour="black")) +
  scale_color_brewer(palette="Set1") +
  remove_x_axis() 

p_middle <- pl_df %>%
  filter(batchEffect=="0") %>%
  ggplot(aes(as.factor(prob_bins), FPR, color=pop, shape=pop)) +
  # geom_violin(scale="width") +
  ggbeeswarm::geom_quasirandom(alpha=0.3, position="stack", size=0.5) +
  geom_point(data = . %>% group_by(method, pop, prob_bins) %>% summarise(FPR=mean(FPR, na.rm = TRUE)), size=2,  
             position=position_dodge(width=0.5)) +
  theme_bw(base_size = 12) +
  scale_color_brewer(palette="Set1", "Population") +
  facet_grid(.~method, labeller=pl_labeller) + 
  geom_hline(yintercept = 0.1, linetype=2) +
  xlab("P(C1) bins") +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5, colour="black"), legend.position = "top", axis.text.y=element_text(colour="black")) 

pl_A <- p_top / p_middle  +
  plot_layout(guides="collect") +
  theme(legend.position="top") 
pl_A

ggsave("fig1dleg.tiff", path = "/Users/kluzer01/Desktop/figures/", height = 4.5, width = 5, device='tiff', dpi=700)
```


```{r, fig.height=6, fig.width=8}
outcome_df <- read_csv("/Users/kluzer01/Documents/milo_analysis_2022/Figure_1/benchmark/data/output_benchmark/outcome_embryo_batchEffect0.csv")
embryo_thresh <- 0.5406022

setwd("/Users/kluzer01/Documents/Milo2.0/R_scripts/milo_analysis_2020/benchmark/data/output_benchmark/")
pl_signal2noise <- function(mypop, myenr, myseed, mymethod){
    if (mymethod=="daseq") {
    pos_outcome <- "PosLFC"
  } else {
    pos_outcome <- "NegLFC"
  }
  popTPR <- outcome_df %>% filter(pop==mypop & enr==myenr & seed==myseed & method==mymethod) %>% pull(TPR)
  res_file <- list.files("/Users/kluzer01/Documents/milo_analysis_2022/Figure_1/benchmark/data/output_benchmark/", 
                           pattern= paste0(".+",'embryo',"_pop_", mypop, "_enr",myenr,"_seed", myseed, '.+batchEffect0',".DAresults.",mymethod,".csv"),
                           full.names = TRUE)
  
  pl <- read_csv(res_file) %>%
    mutate(outcome = case_when(pred=="NotDA" & true_prob < 1-embryo_thresh ~ "FN",
                          pred==pos_outcome & true_prob < 1-embryo_thresh ~ "TP",
                          pred=="NotDA" & true_prob > 1-embryo_thresh ~ "TN"
                          )) %>%
    filter(outcome %in% c('TP', "FN")) %>%
    ggplot(aes(1 - true_prob, color=outcome)) +
    geom_density(size=1) +
    xlab("P(C1)") +
    ggtitle(paste("TPR = ", round(popTPR,2))) +
    theme_bw(base_size = 14)
  pl
}
pl_top <- wrap_plots(lapply(c(0.75, 0.85, 0.95), function(x) pl_signal2noise("Erythroid2", x, myseed = 43, mymethod = "milo_red")), ncol=3)
pl_bot <- wrap_plots(lapply(c(0.75, 0.85, 0.95), function(x) pl_signal2noise("Caudal_neurectoderm", x, myseed = 43, mymethod = "milo_red")), ncol=3)
pl_B <- (pl_top / pl_bot) +
  plot_layout(guides = "collect") 
pl_B
```

